\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed

### 2014-11-03 Paper/grass.tex
Add word count!

\documentclass[aps,prl,twocolumn,superscriptaddress,sort&compress,10pt]{revtex4-1}  % for review and submission
% \documentclass[aps,preprint,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bx}{{\boldsymbol{\hat{x}}}}
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
\newcommand{\ubl}{U_\text{bl}}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

\maketitle
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The hydrodynamic processes resulting from this behavior influence a number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production  \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
One such response of the submerged grass beds to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  
In this letter, we provide a hydrodynamic mechanism for the onset of these coherent oscillations.

Current explanations of \monami ~invoke the existence of a free shear layer at the top of the grass bed (henceforth called grass top) due to vegetation drag  \cite{Ikeda96,Ghisal02,Raupach96}. 
Its instability, through a mechanism similar to the Kelvin-Helmholtz instability, is thought to lead to coherent eddies over the grass bed, and drive large amplitude synchronous oscillations.
This model has been applied to predict the observed frequency of \monami~and to understand transport in the seagrass bed \cite{Nepf00,Ghisal02,Nepf04,Okamoto12}.

However, several aspects of the shear layer model remain unsatisfactory: 
(i)   The velocity profile of the free shear layer is assumed \textit{ad hoc} to be piecewise linear \cite{Delangre06} or hyperbolic tangent \cite{Ghisal02,Raupach96}. 
(ii)  The role of drag in damping the coherent perturbations to the shear profile is sometimes ignored \cite{Raupach96}. 
(iii) No existing theory has explained the threshold flow speed for instability, observed in the lab \cite{Ghisal02} and the field \cite{Grizzle96}, below which \monami ~is not observed.
(iv)  The thickness of the free shear layer is in many cases comparable to the unvegetated layer thickness, and therefore inconsistent with the definition of the free shear layer.
Here we present a mathematical model for the linear instability that accounts correctly for these effects, while also explaining lab experiments and field observations.

Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically, and verify the assumption \textit{a posteriori}.
We show that the drag of the grass results in an instability mode different from the Kelvin-Helmholtz instability.
\begin{figure}
\includegraphics[scale=1]{Grass_Base_Nepf_shear}
% \includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{Schematic setup and comparison of our steady flow profile with that from the experiments by Ghisalberti and Nepf \cite{Nepf04} (Case-I from Table-1 with 1250 plants/m$^2$, 
plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the vegetation, and (iii) a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\Rey \Ndg$ is shown in the inset.}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
% \includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-5mm}
% \includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters estimated from experiments reported by Ghisalberti and Nepf \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from publications by Ghisalberti and Nepf \cite{Ghisal02} and Vivoni \cite{Vivoni98}. 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.}
\label{Re_vs_delta}
\end{figure}

The drag exerted by the vegetation, assumed dense, is modeled by a continuous body force $\mathbf{f}$.
This drag enters the fluid mass and momentum balance equations as 
\begin{equation}
\quad \grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad p+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}
\label{navier-stokes}
\end{equation}
where $\rho$ is the fluid density, $\mathbf{u}$ the velocity, 
$p$ the pressure, $\mu$ the dynamic eddy viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of the grass blade is $O(10^2-10^3)$; therefore, neglecting the skin friction, we model it as the form drag on the vegetation, $\mathbf{f}=N_g C_N \rho u |u| d\bx$ \cite{Nepf99,Nepf00,Nepf04}, where  $N_g$ is the blade number density per unit horizontal area, $C_{N}$ the drag coefficient, and $d$ the average blade width projected perpendicular to the flow. 
In the interest of simplicity, we model the turbulence using an eddy viscosity. 
In the field, both $C_N$, $N_g$ and $\mu$ vary with position, but we do not expect these variations to be central to the instability mechanism, and therefore take them to be constants. 
Based on previous experiments \cite{Vivoni98,Nepf00}, we take $C_N = 1$.

@@

%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
%\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified vegetation number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}
We first calculate the fully developed steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ driven by constant pressure gradient $dP/dx$ as a solution of ~\eqref{navier-stokes}. 
Such a flow satisfies 
\begin{equation}
 -\frac{dP}{dx}+\mu U''(y) +S(y) \rho C_N d N_gU |U|=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the grass bed ($0<y<\hg$) and $S(y)=0$ above it ($\hg< y< 2H$). 
Eq. \eqref{base_equ} is solved subject to no shear at the boundaries, i.e., $U'(0) = U'(2H) = 0$.
The former arises because, in the case of dense vegetation, the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag~\cite{Nepf00}, whereas the latter is due to the free interface. 
This expectation is verified by a comparison, shown in Fig.~\ref{basicflow}, of the steady flow profile from the solution of ~\eqref{base_equ} with experimental measurements.
The profile $U(y)$ has three distinct regions.
Within vegetation, it is approximately uniform with $ U(y) \approx U_g = \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$, arising from the balance between drag and pressure gradient. 
Outside the vegetation, the velocity has a simple parabolic profile due to the balance between viscous forces and the pressure gradient. 
At the grass top, continuity of shear stresses results in a boundary layer of thickness $\delta$. 
Because this boundary layer develops from purely local dynamics, independent of the influence of the boundaries, we identify it to be analogous to the free shear layer \cite{Ghisal02,Nepf04} in the previous explanation of \monami.
Denoting $\ubl$ to be the velocity scale in the boundary layer, and $U_0 = {(dP/dx)~H^2}/{\mu}$ the velocity scale in the unvegetated region, the balance of the viscous forces and the vegetation drag implies $(\mu \ubl/\delta^2 \sim \rho C_N d N_g \ubl^2)$, and the continuity of shear stress across the grass top implies $(\ubl/\delta \sim U_0/H)$,  
Solving for $\delta$ and $\ubl$ yields $\delta/H = \ubl/U_0=(\Rey\Ndg)^{-1/3}$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 
This dependence of the boundary layer thickness on the vegetation density $N_g$, verified in Fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.
The figure also shows that the asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$. In this notation, $U_g/U_0 = (\Rey \Ndg)^{-1/2}$.

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
where the tilde are dropped.
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the submergence ratio of the vegetation $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the mechanism of the instability. 
With these scalings, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \hat\phi(y) \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation \cite{Drazin81} 
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.

A threshold in $\Rey$, above which the flow is unstable (Re$(\sigma)>0$) for at least one $k$, emerges from the 
solution of ~\eqref{Orr-somerfield}. The dependence of this threshold $\Rey$, and the corresponding marginally stable wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown in Fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the vegetation density, indicating a competition between the destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
A similar conclusion was presented \cite{White07} for flow around an emergent (i.e., $\hg>2H$) sea grass patch, but by assuming $U(y)$ to be a tanh-profile, and neglecting the viscous term.
Previous calculations for terrestrial grass differ from our work as they either exclude the vegetation drag in their models \cite{Raupach96}, or assume the mean velocity profile {\it ad hoc} \cite{Raupach96,Delangre06}.
They have not reported a threshold flow condition.

Comparison of the frequency (Im$(\sigma)$) of the fastest growing mode with experimentally observed frequencies in lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in the inset in Fig.~\ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of \monami ~and frequency of vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
The experimentally observed \monami ~wavelengths are not available for comparison.

>>>>

\documentclass[aps,prl,twocolumn,superscriptaddress,sort&compress,10pt]{revtex4-1}  % for review and submission
% \documentclass[aps,preprint,12pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bx}{{\boldsymbol{\hat{x}}}}
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
\newcommand{\ubl}{U_\text{bl}}
\newcommand{\words}[1]{\textbf{(#1)~}}
% \newcommand{\words}[1]{{}}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}t
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

\maketitle
\words{90} Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The hydrodynamic processes resulting from this behavior influence a number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production  \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
One such response of the submerged grass beds to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  
In this letter, we provide a hydrodynamic mechanism for the onset of these coherent oscillations.

\words{80} Current explanations of \monami ~invoke the existence of a free shear layer at the top of the grass bed (henceforth called grass top) due to vegetation drag  \cite{Ikeda96,Ghisal02,Raupach96}. 
Its instability, through a mechanism similar to the Kelvin-Helmholtz instability, is thought to lead to coherent eddies over the grass bed, and drive large amplitude synchronous oscillations.
This model has been applied to predict the observed frequency of \monami~and to understand transport in the seagrass bed \cite{Nepf00,Ghisal02,Nepf04,Okamoto12}.

\words{137} However, several aspects of the shear layer model remain unsatisfactory: 
(i)   The velocity profile of the free shear layer is assumed \textit{ad hoc} to be piecewise linear \cite{Delangre06} or hyperbolic tangent \cite{Ghisal02,Raupach96}. 
(ii)  The role of drag in damping the coherent perturbations to the shear profile is sometimes ignored \cite{Raupach96}. 
(iii) No existing theory has explained the threshold flow speed for instability, observed in the lab \cite{Ghisal02} and the field \cite{Grizzle96}, below which \monami ~is not observed.
(iv)  The thickness of the free shear layer is in many cases comparable to the unvegetated layer thickness, and therefore inconsistent with the definition of the free shear layer.
Here we present a mathematical model for the linear instability that accounts correctly for these effects, while also explaining lab experiments and field observations.

\words{95} Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically, and verify the assumption \textit{a posteriori}.
We show that the drag of the grass results in an instability mode different from the Kelvin-Helmholtz instability.
\begin{figure}
\includegraphics[scale=1]{Grass_Base_Nepf_shear}
% \includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{
\words{129} Schematic setup and comparison of our steady flow profile with that from the experiments by Ghisalberti and Nepf \cite{Nepf04} (Case-I from Table-1 with 1250 plants/m$^2$, 
plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the vegetation, and (iii) a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\Rey \Ndg$ is shown in the inset.
}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
% \includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-5mm}
% \includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{
\words{110} Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters estimated from experiments reported by Ghisalberti and Nepf \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from publications by Ghisalberti and Nepf \cite{Ghisal02} and Vivoni \cite{Vivoni98}. 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.
}
\label{Re_vs_delta}
\end{figure}

\words{28} The drag exerted by the vegetation, assumed dense, is modeled by a continuous body force $\mathbf{f}$.
This drag enters the fluid mass and momentum balance equations as 
\begin{equation}
\quad \grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad p+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}
\label{navier-stokes}
\end{equation}
\words{158} where $\rho$ is the fluid density, $\mathbf{u}$ the velocity, 
$p$ the pressure, $\mu$ the dynamic eddy viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of the grass blade is $O(10^2-10^3)$; therefore, neglecting the skin friction, we model it as the form drag on the vegetation, $\mathbf{f}=N_g C_N \rho u |u| d\bx$ \cite{Nepf99,Nepf00,Nepf04}, where  $N_g$ is the blade number density per unit horizontal area, $C_{N}$ the drag coefficient, and $d$ the average blade width projected perpendicular to the flow. 
In the interest of simplicity, we model the turbulence using an eddy viscosity. 
In the field, both $C_N$, $N_g$ and $\mu$ vary with position, but we do not expect these variations to be central to the instability mechanism, and therefore take them to be constants. 
Based on previous experiments \cite{Vivoni98,Nepf00}, we take $C_N = 1$.

@@

%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
%\end{tabular}
\caption{
\words{69} $Re(\sigma)$ as function of wavenumber and Reynolds number for a specified vegetation number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.
}
\label{K_Re_sigma_set3}
\end{figure*}
\words{32} We first calculate the fully developed steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ driven by constant pressure gradient $dP/dx$ as a solution of ~\eqref{navier-stokes}. 
Such a flow satisfies 
\begin{equation}
 -\frac{dP}{dx}+\mu U''(y) +S(y) \rho C_N d N_gU |U|=0
\label{base_equ}
\end{equation}
\words{377} where $S(y)=1$ in the grass bed ($0<y<\hg$) and $S(y)=0$ above it ($\hg< y< 2H$). 
Eq. \eqref{base_equ} is solved subject to no shear at the boundaries, i.e., $U'(0) = U'(2H) = 0$.
The former arises because, in the case of dense vegetation, the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag~\cite{Nepf00}, whereas the latter is due to the free interface. 
This expectation is verified by a comparison, shown in Fig.~\ref{basicflow}, of the steady flow profile from the solution of ~\eqref{base_equ} with experimental measurements.
The profile $U(y)$ has three distinct regions.
Within vegetation, it is approximately uniform with $ U(y) \approx U_g = \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$, arising from the balance between drag and pressure gradient. 
Outside the vegetation, the velocity has a simple parabolic profile due to the balance between viscous forces and the pressure gradient. 
At the grass top, continuity of shear stresses results in a boundary layer of thickness $\delta$. 
Because this boundary layer develops from purely local dynamics, independent of the influence of the boundaries, we identify it to be analogous to the free shear layer \cite{Ghisal02,Nepf04} in the previous explanation of \monami.
Denoting $\ubl$ to be the velocity scale in the boundary layer, and $U_0 = {(dP/dx)~H^2}/{\mu}$ the velocity scale in the unvegetated region, the balance of the viscous forces and the vegetation drag implies $(\mu \ubl/\delta^2 \sim \rho C_N d N_g \ubl^2)$, and the continuity of shear stress across the grass top implies $(\ubl/\delta \sim U_0/H)$,  
Solving for $\delta$ and $\ubl$ yields $\delta/H = \ubl/U_0=(\Rey\Ndg)^{-1/3}$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 
This dependence of the boundary layer thickness on the vegetation density $N_g$, verified in Fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.
The figure also shows that the asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$. In this notation, $U_g/U_0 = (\Rey \Ndg)^{-1/2}$.

\words{28} Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
\words{124} where the tilde are dropped.
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the submergence ratio of the vegetation $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the mechanism of the instability. 
With these scalings, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \hat\phi(y) \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation \cite{Drazin81} 
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
\words{43} where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.

\words{167} A threshold in $\Rey$, above which the flow is unstable (Re$(\sigma)>0$) for at least one $k$, emerges from the 
solution of ~\eqref{Orr-somerfield}. The dependence of this threshold $\Rey$, and the corresponding marginally stable wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown in Fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the vegetation density, indicating a competition between the destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
A similar conclusion was presented \cite{White07} for flow around an emergent (i.e., $\hg>2H$) sea grass patch, but by assuming $U(y)$ to be a tanh-profile, and neglecting the viscous term.
Previous calculations for terrestrial grass differ from our work as they either exclude the vegetation drag in their models \cite{Raupach96}, or assume the mean velocity profile {\it ad hoc} \cite{Raupach96,Delangre06}.
They have not reported a threshold flow condition.

\words{86} Comparison of the frequency (Im$(\sigma)$) of the fastest growing mode with experimentally observed frequencies in lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in the inset in Fig.~\ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of \monami ~and frequency of vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
The experimentally observed \monami ~wavelengths are not available for comparison.

\begin{table*}
% \footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{\words{17} Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
\words{167} To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

\words{144} The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the top of the vegetation, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood by asymptotically estimating the sizes of the terms in ~\eqref{Orr-somerfield} using $D\sim H/\delta$, $\sigma \sim H/\delta$, and $\ubl \sim \delta/H$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim H/\delta$ (or $\Rey \sim {\Ndg}^{1/2}$) leading to a simplification of ~\eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \sigma \Rey (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
\words{105} in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. 
Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see Fig.~\ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in Fig.~\ref{Asymptotic_mode}, supporting this argument.
